Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-07-11 formalism.tex
Updates from Overleaf

A \emph{graph} $G : \E \rightarrow \Sigma$ is a function from edges to \emph{edge states} $\Sigma = \SetOf{\bot, \Plus, \Minus}$. Edge states determine which edges are live in a graph. For example, if $G(\e) = \Plus$ for some $G : \E \rightarrow \Sigma$, $\e \in \E$,
then we say $\e$ is \emph{live}, i.e., it is part of the live subgraph that is decomposed into a grove, as explained in \autoref{sub:Graph Decomposition}. The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$.

@@

We define the semantics of patch commands via the following transition relation
between graphs.
\[
  G \overset{(s,\e)}{\longrightarrow} G\left[ \e \mapsto s \sqcup G(\e) \right]
\]
Applying patch command $\left(\sigma, \e\right)$ to graph $G$ results in the updated graph $G' = G\left[\e \mapsto \sigma \sqcup G(\e) \right]$, wherein the edge
state associated with edge $\e$ in $G$ becomes the join of $\sigma$ with the state of
$\e$ in $G$, and $G'(\e') = G(\e')$ for any $\e' \ne \e$. Essentially, these patch command semantics act as a transition system between graphs, \emph{joining} the new edge state with the corresponding edge state of the graph to which the patch command is being applied. A \textit{graph patch} is a sequence of graph patch commands. The transition relation is extended to patches as the composition of the relations of the constituent patch commands.

@@

Our source language (\autoref{fig:Syntax}) is based on a simply typed lambda calculus\todo{is there anything here that has to do with the STLC? we just have constructors} with an expression sort and a type sort. Each sort is extended with cases for multi-location references and unicycle references. Additionally, since any position of any constructor may have zero, one, or multiple outgoing edges, we define secondary sorts \textit{ExpList} and \textit{TypList} to represent these possibilities. A source with no outgoing edges corresponds to an empty hole, a source with one outgoing edge corresponds to an ordinary subterm, and a source with multiple outgoing edges corresponds to a local conflict between multiple terms.

% TODO: Should we be more formal and say "is annotated with" instead of "carries"?
These terms also carry information to support additional operations. To enable the reconstruction of the original graph, each constructor carries the UID of its original vertex, each reference carries the vertex that it refers to, and each subterm carries the UID of the edge that leads to it. To support type hole inference (discussion in Sec. TODO), each empty hole and local conflict carries its source. Finally, to support user navigation of the grove, each references also carries the UID of the edge that leads to it.

@@

A live subgraph (or simply graph), which is just a finite set of edges, decomposes into a grove, which is a set of terms. In a tree, each node except the root as a unique parent, and there are no cycles. There is no guarantee of these properties in a graph. Decomposition operates by allowing each `problematic' vertex, in the sense of having multiple parents or forming a cycle, to be the root of its own entry in the grove, and to use references at special term leaves to encode the original graph structure.

@@

\begin{theorem}
    (Classification) TODO
\end{theorem}

@@

>>>>

A \emph{graph} $G : \E \rightarrow \Sigma$ is a function from edges to \emph{edge states} $\Sigma = \SetOf{\bot, \Plus, \Minus}$. Edge states determine which edges are live in a graph. For some $G : \E \rightarrow \Sigma$, $\e \in \E$, $G(\e) = \bot$ if $\e$ has not yet been constructed in the graph. $G(\e) = \Plus$ if $\e$ is \textit{live} in the graph, meaning it is relevant to the user and the type system. $G(\e) = \Minus$ if $\e$ has been deleted. The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$. We assume a graph assigns only finitely many edges to $\Plus$ to ensure the finiteness of the data structure on which further computation is done.

@@

We define the semantics of patch commands via the following transition relation
between graphs.
\[
  G \overset{(s,\e)}{\longrightarrow} G\left[ \e \mapsto s \sqcup G(\e) \right]
\]
Applying patch command $\left(\sigma, \e\right)$ to graph $G$ results in the updated graph $G' = G\left[\e \mapsto \sigma \sqcup G(\e) \right]$, wherein the edge
state associated with edge $\e$ in $G$ becomes the join of $\sigma$ with the state of
$\e$ in $G$, and $G'(\e') = G(\e')$ for any $\e' \ne \e$. Essentially, these patch command semantics act as a transition system between graphs, \emph{joining} the new edge state with the corresponding edge state of the graph to which the patch command is being applied. Since a patch command can only change the value of $G$ on one input, applying the patch command preserves the property that $G$ maps only a finite set of edges to $\Plus$. A \textit{graph patch} is a sequence of graph patch commands. The transition relation is extended to patches as the composition of the relations of the constituent patch commands.

@@

Given a source language parameterized by a set of constructors $k \in \K$ and an arity function $\K \rightarrow \mathbb{N}$, we define an augmented term language as specified in \autoref{fig:Syntax}. Terms are extended with cases for multi-location references and unicycle references. Additionally, since any position of any constructor may have zero, one, or multiple outgoing edges, we define a secondary sort \textit{SubTerm} to represent these possibilities. A source with no outgoing edges corresponds to an empty hole, a source with one outgoing edge corresponds to an ordinary subterm, and a source with multiple outgoing edges corresponds to a local conflict between multiple terms.

% TODO: Should we be more formal and say "is annotated with" instead of "carries"?
These terms also carry information to support additional operations. To enable the reconstruction of the original graph, each constructor carries the UID of its original vertex, each reference carries the vertex that it refers to, and each subterm carries the UID of the edge that leads to it. To support type hole inference (discussion in Sec. TODO), each empty hole and local conflict carries its source. Finally, to support user navigation of the grove, each reference also carries the UID of the edge that leads to it.

@@

A live subgraph, which is just a finite set of edges, decomposes into a set of terms. In a tree, each node except the root as a unique parent, and there are no cycles. There is no guarantee of these properties in a graph. Decomposition operates by allowing each `problematic' vertex, in the sense of having multiple parents or forming a cycle, to be the root of its own entry in the grove, and to use references at special term leaves to encode the original graph structure.

@@

\begin{align}
    \text{parents}_{G}(v) &= \SetOf{v' \in \V \SuchThat{(w, (v', p), v) \in G \land w \in \W \land p \in \P}}
\end{align}

\begin{definition}
    A vertex is an \textit{NP root} in $G$ if $\text{parents}_{G}(v)=\emptyset$
\end{definition}

\begin{definition}
    A vertex is an \textit{MP root} in $G$ if $\abs{\text{parents}_{G}(v)} > 1$
\end{definition}

\begin{definition}
    A vertex is a \textit{U root} in $G$ if $\exists \{v_i\}_{1 \leq i \leq n}$ such that $v_1 = v_n = v$, $\text{parents}_{G}(v_i)=\{v_{i+1}\}$ for all $1 \leq i < n$, and the UID of $v$ is the minimum of the UIDs of $\{v_i\}_{1 \leq i \leq n}$.
\end{definition}

@@

\begin{definition}
    $\text{decomp}_{G}(v) = $
\end{definition}

### 2024-07-11 marking-system.tex
Updates from Overleaf

In \autoref{sub:marking-recap}, we revisit the marked lambda calculus and establish all the required machinery to type-check and mark expressions using a conventional bidirectional system \todo{cite TAPL?}. Next, in \autoref{sub:marking-groves}, we introduce the constructs required to handle the various conflicts that can arise from the graph-based setting when type-checking a tree-native surface language. And finally, we describe the mechanization for the marking system in \autoref{sub:marking-agda} done using the Agda \todo{cite} proof assistant.

@@

First, we introduce here the syntax of the \emph{unmarked} language which comprises two mutually defined expression sorts: $Exp$ \& $ExpList$, denoted by $\EMV$ and $\ELV$ respectively as displayed in \autoref{fig:unmarked-language}. In addition, this unmarked language is an instantiation of the generic term syntax introduced in \autoref{sub:Groves}, with $\EMV$ \& $\TMV$ being the expression and type sorts along with their mutually defined list counterparts $\ELV$ \& $\TLMV$ that allow us to model holes (empty list), \emph{regular} singleton expressions, and local conflicts. Type-checking this base unmarked language is straightforward using conventional bidirectional type-checking rules and we introduce the sort of semantic types $STyp$, denoted by $\sigma$, which builds on ideas from \todo{cite GTLC} Gradually Typed Lambda Calculus (GTLC). This is achieved using 4 mutual type-checking judgments as displayed in \autoref{fig:unmarked-judgments}. The rules are not very different from most simply typed lambda calculi except for the cases where we treat \emph{conflicts} as holes. For the sake of brevity, we have chosen to omit the rules here, but they are expressed in detail in \autoref{sec:marked-unmarked-expressions}.

@@

@@

Next, we introduce the heart of the marking machinery in \autoref{fig:marking-judgments} \todo{add fig}, again in the form of 4 mutually defined judgments along with the marked language denoted by $\ECMV$ \& $\ELMV$. However, here, we add error marks to the expression when one of the premises fails, and type-checking cannot continue further, thereby adding first-class support for error localization. In addition to marking the expressions and type-checking them, at this stage, we also collect inference constraints and \emph{local-info} sets, which come into play when the user has to deal with \emph{multi-location} and \emph{cycle} conflicts.

Now, we examine the marking rules for lambda abstractions and function applications.

\begin{mathpar}
\label{inf:lam-ana}
\inferrule[MKACLam1]{
        \graphErase{\TLMV} = \sigma \\
        \matchedArrowConstraint{\sigma_3}{\sigma_1}{\sigma_2}{\cConstraint_1} \\
        \consistent{\sigma}{\sigma_1} \\
        \anaMarkConstraint{\extendCtx{\ctx}{x}{\sigma}}{\ELV}{\ELMV}{\sigma_2}{\cConstraint_2}{\mConstraint_1}
    }{
        \anaMarkConstraint{\ctx}{\lam{x}{\TLMV}{\EMV}}{\lam{x}{\TLMV}{\ELMV}}{\sigma_3}{\cConstraint_1 \cup \cConstraint_2 \cup \Setof{\constrain{\sigma}{\sigma_1}}}{\mConstraint_1}
    }

@@

@@

>>>>

Finally, we revisit the marked lambda calculus and establish all the required machinery to type-check and mark expressions using a conventional bidirectional system \todo{cite TAPL?}. Next, in \autoref{sub:marking-groves}, we introduce the constructs required to handle the various conflicts that can arise from the graph-based setting when type-checking a tree-native surface language. And finally, we describe the mechanization for the marking system in \autoref{sub:marking-agda} done using the Agda \todo{cite} proof assistant.

@@

\subsubsection{The unmarked language}
\label{sub:unmarked-lang}
First, we introduce here the syntax of the \emph{unmarked} language which comprises two mutually defined expression sorts: $Exp$ \& $ExpList$, denoted by $\EMV$ and $\ELV$ respectively as displayed in \autoref{fig:unmarked-language}. In addition, this unmarked language is an instantiation of the generic term syntax introduced in \autoref{sub:Groves}, with $\EMV$ \& $\TMV$ being the expression and type sorts along with their mutually defined list counterparts $\ELV$ \& $\TLMV$ that allow us to model holes (empty list), \emph{regular} singleton expressions, and local conflicts. Type-checking this base unmarked language is straightforward using conventional bidirectional type-checking rules and we introduce the sort of semantic types $STyp$, denoted by $\sigma$, which builds on ideas from \todo{cite GTLC} Gradually Typed Lambda Calculus (GTLC). This is achieved using 4 mutual type-checking judgments as displayed in \autoref{fig:unmarked-judgments}. The rules are not very different from most simply typed lambda calculi except for the cases where we treat \emph{conflicts} as holes. For the sake of brevity, we have chosen to omit the rules here, but they are expressed in detail in \autoref{sec:marked-unmarked-expressions}.

@@

\subsubsection{The marked language}
\label{sub:marked-lang}
Next, we introduce the heart of the marking machinery in \autoref{fig:marking-judgments} and \autoref{fig:marked-language}, again in the form of 4 mutually defined judgments along with the marked language denoted by $\ECMV$ \& $\ELMV$. However, here, we add error marks to the expression when one of the premises fails, and type-checking cannot continue further, thereby adding first-class support for error localization. In addition to marking the expressions and type-checking them, at this stage, we also collect inference constraints and \emph{local-info} sets, which come into play when the user has to deal with \emph{multi-location} and \emph{cycle} conflicts.

@@

\subsubsection{Marking Lambda Abstractions}
\label{sub:mark-lam}
Now, we examine the marking rules for lambda abstractions and function applications. In \autoref{inf:lam-ana}, in the first case, we successfully mark the lambda abstraction by first \emph{graph-erasing} its type and then using the matched arrow judgment and consistency relations to recursively mark its body. While performing this marking, we gather inference constraints \& local-info maps from the matched arrow judgment and the recursive analysis of the lambda body. Here, all the premises hold, and the abstraction is analyzed with no issues. What happens when the analyzed type is not a matched arrow type or the erased type is \emph{not} consistent with the expected input type? Conventional type-checkers fail here and cannot proceed. This is where the next rules help in localizing these scenarios to the node where the error occurs. In $MKACLam2$, the expected type is not a matched arrow type. So, we mark the abstraction as such using $\ECLamAnaNonMatchedArrow{\var{x}}{\TLMV}{\ELMV}$ which captures the premise that failed while still analyzing the body against the unknown type and gathering constraints. Similarly, in $MKACLam3$, the consistency check between the erased ascription and input of the expected type fails, and again, we mark the abstraction as such, depicting the inconsistency in the ascription of the lambda in $\ECLamInconAsc{\var{x}}{\TLMV}{\ELMV}$.

\begin{mathpar}
\label{inf:mark-lam-ana}
\caption{Marking and Analyzing Lambda Abstractions}
\inferrule[MKACLam1]{
        \graphErase{\TLMV} = \sigma \\
        \matchedArrowConstraint{\sigma_3}{\sigma_1}{\sigma_2}{\cConstraint_1} \\
        \consistent{\sigma}{\sigma_1} \\
        \anaMarkConstraint{\extendCtx{\ctx}{x}{\sigma}}{\ELV}{\ELMV}{\sigma_2}{\cConstraint_2}{\mConstraint_1}
    }{
        \anaMarkConstraint{\ctx}{\lam{x}{\TLMV}{\EMV}}{\lam{x}{\TLMV}{\ELMV}}{\sigma_3}{\cConstraint_1 \cup \cConstraint_2 \cup \Setof{\constrain{\sigma}{\sigma_1}}}{\mConstraint_1}
    }

@@

\subsubsection{Marking Function Applications}
\label{sub:mark-ap-syn}
\begin{mathpar}
\label{inf:ap-syn}
\label{Marking and Synthesizing Function Applications}
\inferrule[MKSCAp1]{
    \synMarkConstraint{\ctx}{\ELV_1}{\ELMV_1}{\sigma}{\cConstraint_1}{\mConstraint_1} \\
    \matchedArrowConstraint{\sigma}{\sigma_1}{\sigma_2}{\cConstraint_2} \\
    \anaMarkConstraint{\ctx}{\ELV_2}{\ELMV_2}{\sigma_1}{\cConstraint_3}{\mConstraint_2} \\
  }{
    \synMarkConstraint{\ctx}{\eApp{u}{\ELV_1}{\ELV_2}}{\eApp{u}{\ELMV_1}{\ELMV_2}}{\sigma_2}{\cConstraint_1 \cup \cConstraint_2 \cup \cConstraint_3}{\mConstraint_1 \cup \mConstraint_2}
  }

  \inferrule[MKSCAp2]{
    \synMarkConstraint{\ctx}{\ELV_1}{\ELMV_1}{\sigma}{\cConstraint_1}{\mConstraint_1} \\
    \notMatchedArrow{\sigma} \\
    \anaMarkConstraint{\ctx}{\ELV_2}{\ELMV_2}{\TUnknown^{\rightarrow_{L}{(exp(u))}}}{\cConstraint_2}{\mConstraint_2}
  }{
    \synMarkConstraint{\ctx}{\eApp{u}{\ELV_1}{\ELV_2}}{\app{\ECApSynNonMatchedArrow{\ELMV_1}}{\ELMV_2}}{\TUnknown^{\rightarrow_{R}{(exp(u))}}}{\cConstraint_1 \cup \cConstraint_2 \cup \SetOf{\constrain{\TUnknown^{exp(u)}}{\tarr{\TUnknown^{\rightarrow_L(exp(u))}}{\TUnknown^{\rightarrow_R(exp(u))}}}}
    }{\mConstraint_1 \cup \mConstraint_2}
  }
\end{mathpar}

Next, in \autoref{inf:ap-syn}, we look at the case when marking an application succeeds. For this to happen, we first synthesize a type for the first sub-term and then match it with an arrow type before analyzing the argument against the expected input type. What if the synthesized is not a matched arrow type? We insert an error mark again, indicating the failed premise in the synthetic mode in $\app{\ECApSynNonMatchedArrow{\ELMV_1}}{\ELMV_2}$.

@@

We observe that \
